---
layout: topic
title: Introduction to R
author: Data Carpentry contributors
minutes: 45
---


```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * Familiarize participants with R syntax
> * Understand the concepts of objects and assignment
> * Understand the concepts of vector and data types
> * Get exposed to a few functions

------------

## Creating objects

```{r, echo=FALSE, purl=TRUE}
### Creating objects in R

```

You can get output from R simply by typing in math in the console

```{r, purl=FALSE}
3 + 5
12/7
```

However, to do useful and interesting things, we need to assign _values_ to
_objects_. To create an object, we need to give it a name followed by the
assignment operator `<-`, and the value we want to give it:

```{r, purl=FALSE}
weight_kg <- 55
```

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. It is also recommended to use nouns for variable names, and verbs for function names.

When assigning a value to an object, R does not print anything. You can print
the value by typing the name:

```{r, purl=FALSE}
weight_kg <- 55
weight_kg
```

Now that R has `weight_kg` in memory, we can do arithmetic with it. For
instance, we may want to convert this weight in pounds (weight in pounds is 2.2
times the weight in kg):

```{r, purl=FALSE}
2.2 * weight_kg
```

We can also change a variable's value by assigning it a new one:

```{r, purl=FALSE}
weight_kg <- 57.5
2.2 * weight_kg
```

This means that assigning a value to one variable does not change the values of
other variables.  For example, let's store the animal's weight in pounds in a new
variable, `weight_lb`:

```{r, purl=FALSE}
weight_lb <- 2.2 * weight_kg
```

and then change `weight_kg` to 100.

```{r, purl=FALSE}
weight_kg <- 100
```

What do you think is the current content of the object `weight_lb`? 126.5 or 200?

## Vectors and data types

```{r, echo=FALSE, purl=TRUE}
### Vectors and data types
```

So far we've stored one value, but typically when doing research we want
to store lots of different values together, like the weights of all of our reserach
organisms. A vector is a group of values, mainly either numbers or
characters. You can assign this list of values to a variable, just like you
would for one item. For example we can create a vector of animal weights:

```{r, purl=FALSE}
weights <- c(50, 60, 65, 82)
weights
```

The `c` stands for "combine", and it combines all of the values (or arguments) into a single vector.

A vector can also contain characters:

```{r, purl=FALSE}
animals <- c("dog", "cat", "sea cumcumber", "sea urchin")
animals
```

To find out how items are in a vector we use `length()`:

```{r, purl=FALSE}
length(animals)
```

The expression `length(...)` is a function call that asks R to run the function `length`.

`length` takes ones argument: the name of the vector who length we want to determine.

The utility of a function is that it will perform its given action on whatever value is passed
to the named argument(s). For example, in this case if we provided the name of a different
vector we get the length of that vector.

```{r, purl=FALSE}
length(weights)
```

There are lots of other we can use on vectors:

```{r, purl=FALSE}
max(weights)
min(weights)
mean(weights)
sum(weights)
```

The function `str()` provides an overview of the object and the elements it
contains. It is a really useful function when working with large and complex
objects:

```{r, purl=FALSE}
str(weights)
str(animals)
```

`str` stands for "structure" and is describinging the structure of the object.

When we used `c()` to originally create the vectors we were also using a function.

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`) and
factors (`factor`).



## Indexing, Sequences, and Subsetting

```{r, echo=FALSE, purl=TRUE}

## Indexing, Sequences, and Subsetting

```

To extract values from a vector we use square brackets. For instance:

```{r, results='show', purl=FALSE}
animals[2]
animals[c(3, 2)]
animals[2:4]
```

If you have experience working in another programming language it's worth noting
that R indexes starting at 1, not 0.

### Conditional subsetting

Besides using the index position of an element in a vector to extract its value
as we saw earlier, we can also use logical vectors:

```{r, purl=FALSE}
animals <- c("mouse", "rat", "dog", "cat")
animals[c(TRUE, FALSE, TRUE, TRUE)]
```

But typically, those logical vectors are not typed by hand but the result of a
logical test:

```{r, purl=FALSE}
animals != "rat"
animals[animals != "rat"]
animals[animals == "cat"]
```

If you can combine multiple tests using `&` (both conditions are true, AND) or `|`
(at least one of the conditions if true, OR):

```{r, purl=FALSE}
animals[animals == "cat" & animals == "rat"] # returns nothing
animals[animals == "cat" | animals == "rat"] # returns both rat and cat
```

If you are trying to combine many conditions, it can become tedious to type. The
function `%in%` allows you to test if a value if found in a vector:

```{r, purl=FALSE}
animals %in% c("rat", "cat", "dog", "duck")
animals[animals %in% c("rat", "cat", "dog", "duck")]
```

In addition to testing equalities, you can also test whether the elements of
your vector are less than or greater than a given value:

```{r, purl=FALSE}
dates <- c(1960, 1963, 1974, 2015, 2016)
dates >= 1974
dates[dates >= 1974]
dates[dates > 1970 & dates <= 2015]
dates[dates < 1975 | dates > 2016]
```

## Missing data

As R was designed to work with data, it includes the concept of missing data.
Missing data are represented as `NA`.

```{r, purl=FALSE}
heights <- c(2, 4, 4, NA, 6)
```

When doing operations on numbers, most functions will return `NA` if the data
you are working with include missing values.

```{r, purl=FALSE}
mean(heights)
max(heights)
```

You can use the optional argument argument `na.rm=TRUE` to calculate the result
while ignoring the missing values. An "optional argument" is one that can either
be specified by the user, or if left out, takes on a *default* value. Optional
arguments are typically used to alter the way the function operates, such as
whether it ignores 'bad values', or what symbol to use in a plot.

```{r, purl=FALSE}
mean(heights, na.rm = TRUE)
max(heights, na.rm = TRUE)
```

Next, we will use the "surveys" dataset to explore the `data.frame` data
structure, which is one of the most common types of R objects.
