---
layout: topic
title: Starting with data
author: Data Carpentry contributors
minutes: 20
---

```{r, echo=FALSE, purl=FALSE, message = FALSE}
source("setup.R")
```

------------

> ## Learning Objectives
>
> * load external data (CSV files) in memory using the survey table
>  (`surveys.csv`) as an example
> * explore the structure and the content of a data frame in R
> * understand what factors are and how to manipulate them

------------

## Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
### Presentation of the survey data
```

We are studying the species and weight of animals caught in plots in our study
area. The dataset is stored as a CSV file: each row holds information for a
single animal, and the columns represent:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxa             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from figshare, and we will use `read.csv()` to
load into memory (as a `data.frame`) the content of the CSV file.

To download the data into the `data/` subdirectory, do:

```{r, eval=FALSE, purl=TRUE}
download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data_joined.csv")
```

You are now ready to load the data:

```{r, eval=TRUE,  purl=FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
```

This statement doesn't produce any output because, as you might recall,
assignment doesn't display anything. If we want to check that our data has been
loaded, we can print the variable's value: `surveys`.

```{r, eval = TRUE, purl = FALSE}
surveys <- read.csv('data/portal_data_joined.csv')
surveys
```

For large datasets like this one that's a lot of output so we often just check the
top few rows of the table using the function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
```

A `data.frame` is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because each column is a
vector, they all contain the same type of data.

```{r, purl=FALSE}
str(surveys)
```

### Challenge

Based on the output of `str(surveys)`, can you answer the following questions?

* What is the class of the object `surveys`?
* How many rows and how many columns are in this object?
* How many species have been recorded during these surveys?

```{r, echo=FALSE, purl=TRUE}

## Challenge
## Based on the output of `str(surveys)`, can you answer the following questions?
## * What is the class of the object `surveys`?
## * How many rows and how many columns are in this object?
## * How many species have been recorded during these surveys?

```

<!---
```{r, echo=FALSE, purl=FALSE}
## Answers
##
## * class: data frame
## * how many rows: 34786,  how many columns: 13
## * how many species: 48
```
--->

As you can see, many of the columns consist of integers, however, the columns
`species` and `sex` are of a special class called a `factor`. Before we learn
more about the `data.frame` class, let's talk about factors. They are very
useful but not necessarily intuitive, and therefore require some attention.
